---
title: "LC50"
author: "A. Proctor and S Wotherspoon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LC50}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Intro



## Model

Lethal concentrations LCx for each combination of additional
stressors are estimated by fitting a Binomial regression model that
relates survival to log toxin concentration, adjusting for background
mortality and assuming a linear relation between log LCx and
descriptors of the additional stressors.

More precisely, the a Binomial regression model parametrized in terms
of log LCx is used to relate survival to log toxin concentration
\[
\begin{align}
	y_{ijk} &\sim \operatorname{Bin}(N_{ijk},\pi_{ijk})\\
	\pi_{ijk} &=
		\begin{cases}
			p_{ijk}q_{k} & C_{ijk} > 0\\
			q_{k} & C_{ijk}= 0
		\end{cases}\\
	\ell_{1}(p_{ijk})&= \alpha_{k} (\log C_{ijk} - \log\mathrm{LCx}_{k}) + z\\
	\log \mathrm{LCx}_{k} &= \beta_{0} + \beta_{1} x_{1k} + \dots + \beta_{m}x_{mk}\\
	\ell_{2}(q_{k}) &= \gamma_{k}
\end{align}
\]

Here \(y_{ijk}\) is the number of survivors at the end of the trial
from \(N_{ijk}\) individuals in replicate \(i\) recieving target toxin
concentration \(j\) and combination \(k\) of additional stressors and
actual measured concentration \(C_{ijk}\), \(\pi_{ijk}\) is the
fraction of individuals expected to survive which is decomposed into a
fraction \(q_{i}\) of individuals expected to survive in the absence
of the toxin, and a fraction \(p_{ijl}\) expected to survive the
additional mortality induced by the toxin.  The fraction \(p_{ij}\) is
related through a monotonic link function \(\ell_{1}\) to the log
measured toxin concentration \(\log C_{ijk}\), the log lethal
concentration \(\log \mathrm{LCx}_{i}\) and a rate parameter
\(\alpha_{i}\) specific to the combination of stressors. Similarly,
the baseline survival fraction \(q_{i}\) is related to a survival
parameter \(\gamma_{i}\) specific to the combination of stressors
through a monotonic link function \(\ell_{2}\).  The lethal concentration
\(\mathrm{LCx}_{i}\) specific to a stressor combination is related
through a log link to a linear predictor constructed from covariates
\(x_{1},x_{2},\ldots,x_{m}\) describing the combination of additional
stressors.  The level of lethality is modelled is determined by the
offset \(z = \ell_{1}(x/100)\), (that is x=50 if LCx=LC50).

The baseline survival parameters \(\gamma_{i}\) and rate parameters
\(\alpha_{i}\) are viewed as nuisance parameters, and inference focuses
on the effects of the additional stressors and the LCx lethal
concentrations.



## Example

The `toxicity` dataset is simulated data showing survival following
exposure to a known toxin in the presence of the additional stressors
temperature and salinity.  In the imagined scenario,  samples of an
aqueous solution of a toxin, of varying concentrations, are prepared
for each of three salinities and are held at one of three constant
temperatures.  Approximately 10 individuals were added to each sample
and the number of survivors recorded at the end of four days exposure.
```{r}
library(LC50)
head(toxicity)
```


### Graphical Exploration

Plotting the survival fraction against concentration for each
combination of additional stressors shows that survival does decrease
with increasing concentrations of the toxin, and in many samples there
is a non-negligible level of control mortality
```{r,fig.width=7,fig.height=6}
library(ggplot2)
ggplot(toxicity,aes(x=conc,y=alive/total,colour=group))+
  geom_point()+
  facet_grid(temperature~salinity)
```


### Model fitting

The LCx lethal concentrations for each temperature and salinity
combination can be estimated with `lcx`.  This function fits the
Binomial model described in the previous section by maximum
likelihood.  The details of the model are specified by three
arguments:

* a formula, the left hand side of which is a two column matrix
  constructed from the number of survivors and mortalities in each
  sample, while the right hand side specifies the terms in the linear
  predictor for log LCx

* the `concentration` argument names the variable in the dataframe
  recording the concentration of the toxin, and

* the `group` argument names a factor in the dataframe that
  disinguishes the combinations of additional stressors.  The model
  requires that the terms of linear predictor for log LCx are
  constant within levels of this factor.

```{r}
fit <- lcx(cbind(alive,dead)~salinity*temperature,concentration=conc,group=group,data=toxicity)
summary(fit)
```
By default, separate background survival rates are estimated for each
group, and the LC50 lethal concentration is modelled using a probit
link.

An analysis of deviance table for the fit is constructed with `anova`.
```{r}
anova(fit,test="Chisq")
```
This table shows no evidence that log LC50 depends on the interaction,
and the model can be simplified to a main effects only model for log
LC50.  Although this fits a restricted model for log LC50, it still
fits a separate control parameter \(\gamma_{i}\) and rate parameter
\(\alpha_{i}\) for each group.
```{r}
fit <- lcx(cbind(alive,dead)~salinity+temperature,concentration=conc,group=group,data=toxicity)
summary(fit)
```

The table of lethal concentrations can be extracted from the summary and displayed graphically
```{r,fig.width=7,fig.height=6}
library(gridExtra)
d.lc50 <- with(summary(fit),
               cbind(lcx,
                     toxicity[match(rownames(lcx),toxicity$group),
                              c("group","salinity","temperature")]))
grid.arrange(
  ggplot(d.lc50,aes(x=group,y=LC50))+
  geom_linerange(mapping=aes(x=group,ymin=`Lwr 95%`,ymax=`Upr 95%`),col="dodgerblue1",lwd=3)+
  geom_point(),
  ggplot(d.lc50,aes(x=salinity,y=LC50,group=temperature,colour=temperature))+
    geom_point()+
    geom_line()+
    geom_errorbar(mapping=aes(ymin=`Lwr 95%`,ymax=`Upr 95%`),width=0.05),
  ggplot(d.lc50,aes(x=temperature,y=LC50,group=salinity,colour=salinity))+
    geom_point()+
    geom_line()+
    geom_errorbar(mapping=aes(ymin=`Lwr 95%`,ymax=`Upr 95%`),width=0.05),
  layout_matrix=rbind(1,2:3))
```

Alternately the **effects** package can be used to construct a graphical summary of the relation between LC50 and the additional stressors.
```{r,fig.width=7,fig.height=6}
library(effects)
plot(Effect(c("salinity","temperature"),fit,
            transformation=list(link=log,inverse=exp)))
```


Predictions from the fitted model can be generated with `predict`.
```{r,fig.width=7,fig.height=6}
d.pr <- expand.grid(temperature=factor(c(12,14,16)),
                    salinity=factor(c(31,33,35)),
                    conc=0:90)
d.pr$group <- with(d.pr,interaction(temperature,salinity,sep="|"))
d.pr$survival <- predict(fit,d.pr)
ggplot(toxicity,aes(x=conc,y=alive/total,colour=group))+
  geom_point()+
  geom_line(aes(x=conc,y=survival,colour=group),data=d.pr)+
  geom_vline(aes(xintercept=LC50,colour=group),data=d.lc50)+
  geom_rect(aes(x=Estimate,y=0,xmin=`Lwr 95%`,xmax=`Upr 95%`,ymin=0,ymax=1,fill=group),data=d.lc50,alpha=0.2,color=NA)+
  facet_grid(temperature~salinity)
```


By default the model is fit with a standard Binomial likelihood.  When there is evidence that the data are overdispersed, a quasi-Binomial model can be fitted by specifying `quasi=TRUE`
```{r}
fitq <- lcx(cbind(alive,dead)~salinity+temperature,concentration=conc,group=group,data=toxicity,quasi=TRUE)
summary(fitq)
```


Alternately, confidence intervals for the lethal concentrations can
be calculated by parametric bootstrap - new data sets are simulated
from the fitted model and the model refitted to the simulated data,
confidence intervals are then derived from the variability seen in the
estimates from the simulated data.
```{r}
d.boot <- toxicity
boot <- lapply(simulate(fit,100),function(s) {
  ## Update data
  d.boot[,colnames(s)] <- s;
  ## Refit model and extract lc50s
  update(fit,data=d.boot,start=fit)})
## Summaries of bootstrap estimates
lc50 <- sapply(boot,function(fit) exp(fit$loglcx))
t(apply(lc50,1,function(lc50) c(LC50=mean(lc50),quantile(lc50,c(0.025,0.975)))))
```


Confidence intervals for predictions can be generated similarly
```{r,fig.width=7,fig.height=6}
pr <- sapply(boot,function(fit) predict(fit,d.pr))
d.pr <- cbind(d.pr,t(apply(pr,1,quantile,c(0.025,0.975))))
ggplot(toxicity,aes(x=conc,y=alive/total,colour=group))+
  geom_point()+
  geom_ribbon(aes(x=conc,y=survival,ymin=`2.5%`,ymax=`97.5%`,fill=group),data=d.pr,alpha=0.2,color=NA)+
  geom_line(aes(x=conc,y=survival,colour=group),data=d.pr)+
  facet_grid(temperature~salinity)
```


### Multiple Comparisons

Multiplicity adjusted pairwise comparison tests can be constructed with
`glht` from the **multcomp** package.
```{r,warning=FALSE}
library(multcomp)
mc <- glht(fit,linfct=mcp(salinity="Tukey",temperature="Tukey"))
summary(mc)
```
Multiplicity adjusted confidence intervals for
differences in log LC50 can be obtained with `confint`
```{r}
confint(mc)
```
These intervals can be converted to confidence intervals for ratios of LC50s by
taking exponentials
```{r}
exp(confint(mc)$confint)
```


### Bayesian Estimates

If **JAGS** and the **rjags** package are installed, Bayesian
estimates of LC50 can be obtained with `lcxJAGS`.  This function is
an analog of `lcx` that constructs an object of class `jags` that can
be used to sample from the posterior for the model using the
facilities of the **rjags** package.

To generate Bayesian estimates for the reduced model, `lcxJAGS` is
used to generate a model object
```{r}
library(rjags)
model <- lcxJAGS(cbind(alive,dead)~salinity+temperature,concentration=conc,group=group,data=toxicity)
```
and an initial burn-in sample is drawn
```{r}
update(model,1000)
```
A final sample is then drawn and stored.  It is possible to sample for
any of the model variables "alpha", "beta", "gamma", "lcx",
"loglcx", "p" or "q" as required.
```{r}
s <- coda.samples(model,c("beta","q","lcx"),n.iter=10000,n.thin=10)
summary(s)
```

### Shrinkage

In trials with few distinct concentrations, the rate parameters
\(\alpha_{i}\) can be poorly resolved.  When these rates are
overestimated the confidence intervals for the LC50 can be too
short. The `rate.shrink` parameter adds a penalty to the likelhood to
shrink the \(\alpha_{i}\) back towards zero.

Without shrinkage, the estimates of some of the rate parameters for
the toxicity data are particularly high, and the confidence intervals
for the corresponding LC50s are not very wide.
```{r}
fit <- lcx(cbind(alive,dead)~salinity+temperature,concentration=conc,group=group,data=toxicity)
summary(fit,rate=TRUE)
```

```{r,fig.width=7,fig.height=6}
d.pr <- expand.grid(temperature=factor(c(12,14,16)),
                    salinity=factor(c(31,33,35)),
                    conc=0:90)
d.pr$group <- with(d.pr,interaction(temperature,salinity,sep="|"))
d.pr$survival <- predict(fit,d.pr)
ggplot(toxicity,aes(x=conc,y=alive/total,colour=group))+
  geom_point()+
  geom_line(aes(x=conc,y=survival,colour=group),data=d.pr)+
  #geom_vline(aes(xintercept=LC50,colour=group),data=d.lc50)+
  #geom_rect(aes(x=Estimate,y=0,xmin=`Lwr 95%`,xmax=`Upr 95%`,ymin=0,ymax=1,fill=group),data=d.lc50,alpha=0.2,color=NA)+
  facet_grid(temperature~salinity)
```

With shrinkage,
```{r}
fit <- lcx(cbind(alive,dead)~salinity+temperature,concentration=conc,group=group,data=toxicity,rate.shrink=0.001)
summary(fit,rate=TRUE)
```

```{r,fig.width=7,fig.height=6}
d.pr <- expand.grid(temperature=factor(c(12,14,16)),
                    salinity=factor(c(31,33,35)),
                    conc=0:90)
d.pr$group <- with(d.pr,interaction(temperature,salinity,sep="|"))
d.pr$survival <- predict(fit,d.pr)
ggplot(toxicity,aes(x=conc,y=alive/total,colour=group))+
  geom_point()+
  geom_line(aes(x=conc,y=survival,colour=group),data=d.pr)+
  #geom_vline(aes(xintercept=LC50,colour=group),data=d.lc50)+
  #geom_rect(aes(x=Estimate,y=0,xmin=`Lwr 95%`,xmax=`Upr 95%`,ymin=0,ymax=1,fill=group),data=d.lc50,alpha=0.2,color=NA)+
  facet_grid(temperature~salinity)
```


## LCx

By default all calculations are performed based on LC50, but it is
possible to perform the analysis based on other lethal concentrations.

To repeat the basic analysis above for LC10
```{r}
fit <- lcx(cbind(alive,dead)~salinity*temperature,concentration=conc,group=group,data=toxicity,lethal=10)
summary(fit)
```

Changing the reference lethality imacts inferences - for the LC10
there is strong evidence of a salinity by temperature interaction
```{r}
anova(fit,test="Chisq")
```

```{r,fig.width=7,fig.height=6}
d.lcx <- with(summary(fit),
               cbind(lcx,
                     toxicity[match(rownames(lcx),toxicity$group),
                              c("group","salinity","temperature")]))
ggplot(d.lcx,aes(x=group,y=LC10))+
  geom_linerange(mapping=aes(x=group,ymin=`Lwr 95%`,ymax=`Upr 95%`),col="dodgerblue1",lwd=3)+
  geom_point()
```




```{r,fig.width=7,fig.height=6}
d.pr <- expand.grid(temperature=factor(c(12,14,16)),
                    salinity=factor(c(31,33,35)),
                    conc=0:90)
d.pr$group <- with(d.pr,interaction(temperature,salinity,sep="|"))
d.pr$survival <- predict(fit,d.pr)
ggplot(toxicity,aes(x=conc,y=alive/total,colour=group))+
  geom_point()+
  geom_line(aes(x=conc,y=survival,colour=group),data=d.pr)+
  geom_vline(aes(xintercept=LC5,colour=group),data=d.lcx)+
  geom_rect(aes(x=Estimate,y=0,xmin=`Lwr 95%`,xmax=`Upr 95%`,ymin=0,ymax=1,fill=group),data=d.lcx,alpha=0.2,color=NA)+
  facet_grid(temperature~salinity)
```


